---
title: "FitBit fitness tracker data analysis"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup_1, include=FALSE}
# deactivate any messages and warnings to show
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Introduction {#intro}
This file contains analysis and visualizing of 
[FitBit Fitness Tracker Data](https://www.kaggle.com/datasets/arashnic/fitbit) as a part of case study
from [Google Data Analytics Professional Certificate program](https://www.coursera.org/professional-certificates/google-data-analytics).
The main purpose of this analysis is searching for any trends and insights, creating
meaningful visualizations to generate ideas for improving business and user experience. 
Chapters of the file:

1. [Introduction](#intro)
2. [Setup](#setup)
3. [Importing data](#import)
4. [Data overview](#overview)
4. [Analysis](#analysis)
5. [Conclusions](#conclusions)


## Setup {#setup}
First of all, we need to load all required libraries we will use to analyze and
visualize data.
```{r setup_2}
library(tidyverse)
library(ggalt)
library(cowplot)
```


## Importing data {#import}
To start working with data we need to import all datasets we will use for analysis
(all datasets were created using SQL on step "Process" of case study project). 
We won't use data about weight and heart rate due to small user samples of 
these datasets.

```{r importing_datasets}
daily_activity_total <- read.csv('prepared_data/daily_activity_total.csv')

hourly_intensity_total <- read.csv('prepared_data/hourly_intensity_total.csv')

hourly_steps_total <- read.csv('prepared_data/hourly_steps_total.csv')

hourly_calories_total <- read.csv('prepared_data/hourly_calories_total.csv')

daily_sleep_total <- read.csv('prepared_data/daily_sleep_total.csv')

average_sleep_by_weekdays <- read.csv('prepared_data/average_sleep_by_weekdays.csv')

bad_sleep_total <- read.csv('prepared_data/bad_sleep_total.csv')
```


## Data overview {#overview}
Before starting the analysis, let's look at the datasets we have.

```{r daily_activity_total_overview}
head(daily_activity_total, 10)
```

```{r hourly_intensity_total_overview}
head(hourly_intensity_total, 10)
```

```{r hourly_steps_total_overview}
head(hourly_steps_total, 10)
```

```{r hourly_calories_total_overview}
head(hourly_calories_total, 10)
```

```{r daily_sleep_total_overview}
head(daily_sleep_total, 10)
```

```{r average_sleep_by_weekdays_overview}
head(average_sleep_by_weekdays, 10)
```

```{r bad_sleep_total_overview}
head(bad_sleep_total, 10)
```


## Analysis {#analysis}
Now we can start our main part of analysis. We will research two aspects of
user data: user activity and user sleep.

Let's start with exploring number of steps users walk everyday:

```{r daily_steps_vs_calories}
# create scatter plot with trend line for daily activity
daily_activity_total %>%
  ggplot(mapping = aes(x = TotalSteps, y = Calories, colour = period)) +
  geom_point(size = 1) + 
  geom_smooth(mapping = aes(linetype = ''), 
              se = FALSE, color = '#ff5959') +
  scale_color_manual(values = c('march-april' = '#57abff', 
                                'april-may' = '#304bff'), 
                     name = 'Periods') + 
  scale_linetype(name = "Trend line") + theme_bw() +
  labs(x = 'Steps',
       title = 'Relationship between number of steps and burned calories per day')
```

We can see strong relationship between number of steps and 
number of burned calories per day. This fact is pretty obvious but now 
we can see it clearly. And what about this relationship during a day by hours?

```{r hourly_steps_vs_calories}
# create line plot for steps
steps_plot <- hourly_steps_total %>%
  # group by hour of day and period
  group_by(time_hour = hour(ActivityHour), period) %>% 
  summarise(steps = mean(StepTotal)) %>%
  ggplot(mapping = 
           aes(x = time_hour, y = steps, group = period, 
              colour = period)) + 
  geom_point() + geom_xspline() +
  scale_color_manual(values = c('march-april' = '#57abff', 
                                'april-may' = '#304bff'), name = 'Periods') + 
  theme_bw() +
  labs(x = NULL, y = NULL,
       title = '',
       subtitle = 'Average number of steps during a day')


# create line plot for calories
calories_plot <- hourly_calories_total %>%
  # group by hour of day and period
  group_by(time_hour = hour(ActivityHour), period) %>%
  summarise(calories = mean(Calories)) %>%
  ggplot(mapping = aes(x = time_hour, y = calories, 
                       group = period, color = period)) + 
  geom_point() + geom_xspline() +
  scale_color_manual(values = c('march-april' = '#57abff', 
                                'april-may' = '#304bff'), name = 'Periods') +
  theme_bw() +
  labs(x = NULL, y = NULL, 
       subtitle = 'Average number of burned calories during a day')


# display both charts together to compare them
plot_grid(steps_plot, calories_plot, nrow = 2)
```

And here we can see how similar these patterns are. So thanks to these charts we 
made sure that walking is important to burn calories.

But what about real number of steps users walk everyday in average?
```{r average_steps_by_user}
# group records by user Id to find average number of steps of every user
daily_activity_grouped <- daily_activity_total %>%
  group_by(Id) %>%
  summarise(average_steps = mean(TotalSteps))

# create bar plot for average number of steps of every user
daily_activity_grouped %>%
  ggplot(mapping = aes(x = reorder(as.character(Id), average_steps), 
                       y = average_steps, 
                       fill = average_steps > 6000)) +
  geom_bar(stat = 'identity', position = 'dodge') +
  scale_fill_manual(values = c('FALSE' = '#ff5959', 'TRUE' = '#304bff'), 
                    name = 'Steps > 6000') +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank()) + 
  labs(x = NULL, y = 'Average number of steps',
       title = "Average number of steps each user walk per day")

# total number of users
length(daily_activity_grouped$average_steps)
# number of users who walk less than 6000 steps per day in average
sum(daily_activity_grouped$average_steps < 6000)
```

We can see that 14 of 35 users walk less than 6000 steps per day in average
(when it is recommended to walk at least 6000-8000 steps per day).

Next, let's review user's activity at the whole. We will start with time distributions 
of different activity types by days of week:
```{r daily_intensity_types}
# create bar plot for every intensity type
daily_activity_total %>%
  group_by(week_day = wday(ActivityDate, label = TRUE, abbr = FALSE, 
                           week_start = 1, locale = 'English')) %>%
  # summarize by every type of intensity
  summarise(very_active = mean(VeryActiveMinutes),
            fairly_active = mean(FairlyActiveMinutes),
            lightly_active = mean(LightlyActiveMinutes)) %>%
  pivot_longer(cols = c('very_active', 'fairly_active', 'lightly_active'),
               names_to = 'intensity', values_to = 'minutes') %>%
  transform(
    # order intensity types from top to bottom within legend
    intensity = factor(intensity, 
                       levels = c('lightly_active',
                                  'fairly_active',
                                  'very_active'),
                       labels = c('Lightly active',
                                  'Fairly active',
                                  'Very active'))) %>%
  ggplot(mapping = aes(x = week_day, y = minutes, fill = intensity)) + 
  geom_bar(stat = 'identity', position = 'dodge') + 
  facet_grid(rows = vars(intensity), scales = 'free') + 
  scale_fill_manual(values = c('Lightly active' = '#52c5ff',
                               'Fairly active' = '#5294ff',
                               'Very active' = '#0f73ff'), name = 'Intensities') + 
  theme_bw() + 
  labs(x = NULL, y = 'Minutes',
       title = 'Time of different intensity types')
```

As we can see, average time of lightly active intensity per day is around 180-190 minutes,
average time of fairly active intensity is around 13-14 minutes and time of very active time
is around 19-20 minutes.

Let's research relationships between intensity time and other statistics:

```{r hourly_calories_vs_intensity}
# merge calories dataset and intensity dataset to
# create scatter plot with trend line
merge(x = hourly_calories_total, 
      y = hourly_intensity_total, 
      by.x = c('Id', 'ActivityHour', 'period'),
      by.y = c('Id', 'ActivityHour', 'period')) %>%
  ggplot(mapping = aes(x = Calories, y = TotalIntensity, colour = period)) +
  geom_point(size = 1) + 
  geom_smooth(mapping = aes(linetype = ''), se = FALSE, color = '#ff5959') +
  scale_color_manual(values = c('march-april' = '#57abff','april-may' = '#304bff'),
                     name = 'Periods') + 
  scale_linetype(name = "Trend line") + theme_bw() +
  labs(x = 'Calories', y = 'Intensity value',
       title = 'Relationship between burned calories and intensity value')
```

Quite obvious that there is positive correlation between intensity time and
number of burned calories per hour.

```{r daily_sleep_vs_intensity}
# merge daily sleep dataset and hourly intensity dataset together
# to create scatter plot with trend line
merge(x = daily_sleep_total %>% transform(SleepDay = date(SleepDay)),
      y = hourly_intensity_total %>% 
        group_by(Id, activity_day = date(ActivityHour), period) %>%
        summarise(total_intensity = sum(TotalIntensity)),
      by.x = c('Id', 'SleepDay', 'period'),
      by.y = c('Id', 'activity_day', 'period')) %>%
  ggplot(mapping = aes(x = TotalMinutesAsleep, y = total_intensity, 
                       colour = period, group = 1)) +
  geom_point(size = 1) +
  geom_smooth(mapping = aes(linetype = ''), 
              se = FALSE, color = '#ff5959') +
  scale_color_manual(values = c('march-april' = '#57abff', 
                                'april-may' = '#304bff'), 
                     name = 'Periods') + 
  scale_linetype(name = "Trend line") + theme_bw() +
  labs(x = 'Minutes asleep',
       y = 'Intensity value',
       title = 'Relationship between sleep time and intensity value')
```

Not so obvious: there is no correlation between intensity time and sleep time. 


```{r daily_sleep_vs_sedentary_time}
# merge daily sleep dataset and daily activity dataset together 
# to create scatter plot with trend line
merge(x = daily_sleep_total,
      y = daily_activity_total,
      by.x = c('Id', 'SleepDay', 'period'),
      by.y = c('Id', 'ActivityDate', 'period')) %>%
  ggplot(mapping = aes(x = SedentaryMinutes, y = TotalMinutesAsleep, colour = period)) +
  geom_point(size = 1) +
  geom_smooth(mapping = aes(linetype = ''), 
              se = FALSE, color = '#ff5959') +
  scale_color_manual(values = c('march-april' = '#57abff', 
                                'april-may' = '#304bff'), 
                     name = 'Periods') + 
  scale_linetype(name = "Trend line") + theme_bw() +
  labs(x = 'Sedentary minutes',
       y = 'Minutes asleep',
       title = 'Relationship between sleep time and sedentary time, minutes')
```

And here we can see an interesting trend: there is negative correlation
between sleep time and sedentary time. Thus we can say that spending less time
sitting can improve user's sleep.

But how users really sleep? Let's answer this question:

```{r weekdays_average_sleep_time}
# create bar plot for sleep time by days of week
average_sleep_by_weekdays %>% 
  transform(WeekDay = wday(1:7, label = TRUE, 
                           abbr = FALSE, locale = 'English')) %>%
  ggplot(mapping = aes(x = WeekDay, y = AverageMinutesAsleep)) +
  geom_bar(stat = 'identity', fill = '#4f85ff') + theme_bw() +
  labs(x = NULL, y = 'Minutes',
       title = 'Average sleep time, minutes')
```

In average users sleep longer on weekends and most of the week they sleep less
than 7 hours (which is minimum recommended sleep time).

```{r average_sleep_time_by_user}
# group records by user Id to find average sleep time of every user
daily_sleep_grouped <- daily_sleep_total %>%
  group_by(Id) %>%
  summarise(average_sleep_minutes = mean(TotalMinutesAsleep))

# create bar plot for average sleep time by user
daily_sleep_grouped %>%
  ggplot(mapping = aes(x = reorder(as.character(Id), average_sleep_minutes), 
                       y = average_sleep_minutes, 
                       fill = average_sleep_minutes >= 420)) +
  geom_bar(stat = 'identity', position = 'dodge') +
  scale_fill_manual(values = c('FALSE' = '#ff5959', 'TRUE' = '#304bff'), 
                    name = 'Sleep time, minutes >= 420') + theme_bw() +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank()) + 
  labs(x = NULL, y = 'Average sleep time',
       title = "Average sleep time of every user, minutes")

# total number of users
length(daily_sleep_grouped$average_sleep_minutes)
# number of users who sleep less than 7 hours per day in average
sum(daily_sleep_grouped$average_sleep_minutes < 420)
```

Most of the users sleep less than 7 hours in average! Some of them even sleep
less than 3 hours! This requires special attention but we need more data 
for deeper analysis.

```{r bad_sleep}
# create bar plot for number of bad sleep days 
# (when user sleeps less than 7 hours) by days of week
merge(bad_sleep_total,
      bad_sleep_total %>% group_by(Id) %>% summarise(total_days = sum(days)), 
      by = c('Id')) %>%
  ggplot(mapping = aes(x = reorder(as.character(Id), -total_days), 
                       y = days, fill = period)) + 
  geom_bar(stat = 'identity') +
  scale_fill_manual(values = c('march-april' = '#59a7ff', 
                               'april-may' = '#4278ff'), 
                    labels = c('march-april', 'april-may'),
                    name = 'Periods') + theme_bw() + 
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank()) +
  labs(x = NULL, y = 'Number of days', 
       title = "Number of days each user slept less than 7 hours")
```

And here we can see that some of the users most of the days slept less 
than 7 hours per day. Lack of sleep can be serious problem so this situation 
requires special attention.


## Conclusions {#conclusions}
We have analysed data from users fitness trackers for two months, 
have researched users physical activity and sleep and now we can summarise gained results:

* walking and physical activity is important to burn more calories per day
* the most active period of day is from 10 to 18 hours
* more than 1/3 of users walk less than 6000 steps which is the minimal recommended number of steps
* more than 80% of time users spend sedentary
* 2/3 of users sleep less than 7 hours per day in average
* decreasing sedentary time can lead to better sleep